---
title: "Aging Indicator Analysis"
author: "Don Boyd"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_notebook:
    df_print: paged
    toc: yes
    toc_depth: 5
always_allow_html: yes    
    
editor_options: 
  chunk_output_type: console
---

<!-- comments -->

# Notes: Code folding, etc.
<!--
# code folding ----
#   alt-L, alt-shift-L  one section
#   alt-O, alt-shift-O  all sections
#   alt-R run current code section (remapped from Ctrl-alt-T)

# ctrl-D run current chunk
# ctrl-R run current selection
-->


# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, eval=FALSE)
```

```{r system_specific_values, include=FALSE}

```

```{r globals, include=FALSE}
# 11 states with diverse economies and tax structures
sts <- c("CA", "CT", "FL", "IL", "NJ", "NY", "PA", "TN", "TX", "WA", "WY")

# colors lighter to darker
# #fee0d2 #fc9272 #de2d26
red2 <- c("#fee0d2", "#de2d26")
# #deebf7 #9ecae1 #3182bd
blue2 <- c("#deebf7", "#3182bd")
# #e5f5e0 #a1d99b #31a354
green2 <- c("#e5f5e0", "#31a354")

```

```{r libraries, include=FALSE}
library("magrittr")
library("plyr") # needed for ldply; must be loaded BEFORE dplyr
library("tidyverse")
options(tibble.print_max = 60, tibble.print_min = 60) # if more than 60 rows, print 60 - enough for states
# ggplot2 tibble tidyr readr purrr dplyr stringr forcats
library("readxl")

library("scales")
library("hms") # hms, for times
library("lubridate") # lubridate, for date/times
library("vctrs")

library("grDevices")
library("knitr")
library("kableExtra")

library("DBI")

library("btools")
library("bdata")

library("BEAData")
library("qtax")

library("DT") # for datatable

library("zoo") # for rollapply

library("broom") # for automating the cleanup of complex output

library("quantmod")
library("forecast")
library("tsoutliers")
library("timetk")

library("janitor")

library("ggrepel")

library("maps")
library("usmap")
library("gridExtra")

```

```{r functions, include=FALSE}
ma_db <- function(x, period) {
  # create trailing moving average of x of length period
  zoo::rollapply(x, period, function(x) mean(x, na.rm=TRUE), fill=NA, align="right")
}

stname <- function(stabbr) {state.name[match(stabbr, state.abb)]}
stname(sts) # sts defined above

get_stabbr <- function(x){bdata::stcodes$stabbr[match(x, bdata::stcodes$stfips)]}

```

```{r map_setup, include=FALSE}
#.. Functions
theme_map <- function(base_size=9, base_family="") {
  # see:
  # https://socviz.co/maps.html
  # https://github.com/kjhealy/socviz
  # https://cran.r-project.org/web/packages/usmap/vignettes/mapping.html
  require(grid)
  theme_bw(base_size=base_size, base_family=base_family) %+replace%
    theme(axis.line=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank(),
          axis.title=element_blank(),
          panel.background=element_blank(),
          panel.border=element_blank(),
          panel.grid=element_blank(),
          panel.spacing=unit(0, "lines"),
          plot.background=element_blank(),
          legend.justification = c(0, 0),
          legend.position = c(0, 0)
    )
}


statemap <- function(mapdf, mapvar, cutpts, cutlabels, map_colors, map_title, legend_title){
  mapdf <- mapdf %>%
    rename(mapvalue=mapvar) %>%
    mutate(data_group=cut(mapvalue, cutpts, labels=cutlabels),
           mapvarname=mapvar) %>%
    select(stabbr, mapvarname, mapvalue, data_group)
  
  mdata <- left_join(
    usmap::us_map() %>% arrange(full, piece, order), 
    mapdf %>% rename(abbr=stabbr),
    by="abbr")
  
  p <- mdata %>%
    ggplot(aes(x = x, y = y)) +
    geom_polygon(aes(fill=data_group, group = group),
               color = "gray90", size = 0.1) +
    scale_fill_manual(values=map_colors, drop=TRUE, na.translate=FALSE) + # na.translate drops NA from the legend
    coord_equal() + 
    guides(fill=guide_legend(title=legend_title)) +
    ggtitle(map_title) +
    geom_text(data = bdata::statemap_labels %>% filter(stabbr != "DC"), aes(x, y, label = stabbr), size = 3) +
    theme_map() +
    theme(legend.position = "right") +
    theme(plot.title = element_text(size = 14, face = "bold"))
  
  #   labs(caption=capt) +
  #   theme(plot.caption = element_text(hjust=0, size=8))
  return(p)
}

```


# Introduction
Goals for the indicator:

* Measure selected fiscal and/or economic effects related to changes in the age structure of states' populations
* Help policymakers understand how these demographic changes will affect states' fiscal health
* Regularly updated
* Can be fiscal or demographic; but goal is to understand impacts on fiscal health

Considerations:

* Should it be fiscal or demographic?
* What would an indicator concentrated on spending pressures look like?
* Should the aging indicator present projections?
* Is there a value to following PRC?? approach for the new indicator?  Would it even be possible to use Census data for state-level projections?
* What are some common issues or pitfalls in aging-related research that we should think about as we develop this indicator?

Barb asks:

* Would changing size of prime-age workers by state over time get at effects of aging on a state's economic prospects?


# Data preparation
## ONETIME
```{r ONETIME_pop_age}
# https://factfinder.census.gov/bkmk/table/1.0/en/PEP/2018/PEPSYASEX?
# https://www2.census.gov/programs-surveys/popest/datasets/2010-2019/
# PEP_2018_PEPSYASEX_with_ann.csv # annotations merged
fn <- "PEP_2018_PEPSYASEX_with_ann.csv"
vnames <- read_csv(here::here("data/popests/PEP_2018_PEPSYASEX/", fn), n_max = 0) %>% names
df <- read_csv(here::here("data/popests/PEP_2018_PEPSYASEX/", fn), col_names = vnames, skip=2) # skip 2 so that we just read data
ns(df) %>% ht
df[1:4, 1:5]
dim(df)

# long file
get_stabbr <- function(x){bdata::stcodes$stabbr[match(x, bdata::stcodes$stfips)]}
# get_stabbr(c("01", "02", "36"))
dfl <- df %>%
  rename(stname=`GEO.display-label`,
         stfips=GEO.id2) %>%
  select(-GEO.id) %>%
  pivot_longer(-c(stname, stfips)) %>%
  mutate(stfips=ifelse(stname=="United States", "00", stfips),
         stabbr=get_stabbr(stfips))
glimpse(dfl)
ht(dfl)
count(dfl, stabbr, stname, stfips)

# https://regexone.com/
# regex101.com

# dfl2 <- dfl %>%
#   mutate(src=str_extract(name, "(?:(?!sex).)*"))
#  mutate(src=str_split(name, "sex", simplify = TRUE)[, 1])

dfl2 <- dfl  %>%
  separate(col=name, into=c("src", "part2"), sep="(sex)", remove=FALSE) %>%
  separate(col=part2, into=c("sex", "part3"), remove=FALSE) %>%
  mutate(vtype=ifelse(part3 != "medage", "pop", "medage")) %>%
  separate(col=part3, into=c("part4", "age"), sep="age", remove=FALSE) %>%
  mutate(age=ifelse(age=="85plus", "85", age),
         age=as.integer(age),
         sex=as.integer(sex),
         year=str_sub(src, -4, -1) %>% as.integer,
         month=str_sub(src, -5, -5) %>% as.integer,
         src=str_sub(src, 1, 3))

glimpse(dfl2)
ht(dfl2)
count(dfl2, age) %>% ht # 85 is 85plus, 999 is total
count(dfl2, sex) # 0 is total
count(dfl2, year)
count(dfl2, month)
count(dfl2, src)
count(dfl2, vtype)

dfl3 <- dfl2 %>%
  select(stabbr, src, year, month, vtype, sex, age, value)
glimpse(dfl3)
saveRDS(dfl3, file=here::here("data", "popagesex.rds"))

```

```{r ONETIME_USMDB_life}
#.. USMDB documentation ----
# The USMDB death rates and life tables have been computed as in the Human Mortality Database, following six steps, corresponding to each of six data types.
# Here is an overview of the process:

# Births. Annual counts of live births by sex have been collected for each population for all years since 1959. 
#   These counts have been used mainly for making population estimates at younger ages.
# Deaths. Death tabulations were built from the NCHS Mortality Files at the finest level of detail available and 
#   uniform methods were implemented to estimate death counts by completed age (i.e., age-last-birthday at time of death), calendar year
#   of death, and calendar year of birth.
# Population size. July 1st annual estimates of population size were obtained from the US Census Bureau for years since 1970 and inter-censal
#   population estimates were constructed for years 1959-1969 from a combination of US Census data, birth and death counts.
# Exposure-to-risk. Estimates of the population exposed to the risk of death during each age-time interval were estimated
#   from the annual population estimates, after correcting for the timing of deaths within each interval.
# Death rates. Death rates were computed as the ratios of the death counts for a given age-time interval divided by the estimate
#   of the exposure-to-risk in the same interval.
# Life tables. Probabilities of death were computed from death rates and used to construct all other life table functions,
#   including life expectancies and other useful indicators of mortality and longevity. 

# The life tables include the following columns
# Year     Calendar year or range of years of occurrence
# Age      Age group for n-year interval from exact age x to just before exact age x+n
# m(x)     Central death rate between age x and age x+n
# q(x)     Probability of death between age x and age x+n
# a(x)     Average length of survival between age x and age x+n for persons dying in the interval
# l(x)     Number of survivors at exact age x, assuming l(0) = 100,000
# d(x)     Number of deaths between age x and age x+n
# L(x)     Number of person-years lived between age x and age x+n
# T(x)     Number of person-years remaining after exact age x
# e(x)     Life expectancy at exact age x (in years) = remaining length of life for survivors to age x



#.. get USMDB ----
ltdir <- here::here("data/lifetables/States")
df <- list.files(path=ltdir, pattern="*_bltper_1x1.csv", full.names=TRUE, recursive=TRUE) %>% 
  map_df(~read_csv(.))
glimpse(df)
# popname, sex, year, age, mx, qx, ax, lx, dx, Lx, Tx, ex

df2 <- df %>%
  rename(stabbr=PopName, sex=Sex, year=Year, age=Age) %>%
  mutate(age=ifelse(age=="110+", "110", age),
         age=as.integer(age))
glimpse(df2)
df2

df2 %>%
  filter(stabbr=="NY", year==2010, age<110) %>%
  ggplot(aes(x=mx, y=qx)) + geom_point() +
  geom_abline(slope=1)

df2 %>%
  filter(age==atage) %>%
  filter(stabbr %in% sts) %>%
  ggplot(aes(year, mx, colour=stabbr)) +
  geom_line() +
  geom_point() +
  ggtitle(paste0("Death rate at age ", atage, ", selected states")) +
  scale_x_continuous(name=NULL, breaks=seq(1950, 2030, 5)) +
  scale_y_continuous(name="Death rate", breaks=seq(0, 0.1, .002)) +
  theme_bw()

saveRDS(df2, file=here::here("data", "usmdb.rds"))

```

```{r ONETIME_ACS_disability}
# ACS disability ----
# https://www.census.gov/programs-surveys/acs/technical-documentation/pums/documentation.html

dbdir <- "D:/Data/CensusACS/RSQLite1YearDB/"
dbf <- paste0(dbdir, "acs_1year.sqlite")

con <- DBI::dbConnect(RSQLite::SQLite(), dbf)

tname <- "p2018"

dbListTables(con)
dbListFields(con, tname)

selstring <- paste0("SELECT * FROM ", tname, " LIMIT 5")
tmp <- dbGetQuery(con, selstring)
glimpse(tmp)

selstring <- paste0("SELECT COUNT(*) FROM ", tname)
dbGetQuery(con, selstring )

# DIS Character 1
# Disability recode
# 1 .With a disability
# 2 .Without a disability

# institutionalized group quarters relp==16

# vars <- c("st", "pwgtp", "agep", "dis")
# vars <- expression(st, pwgtp, agep, dis)
# bquote(st, pwgtp, agep, dis)
# paste0("a", vars)

(selstring <- paste0("SELECT ", vars, " FROM ", tname, " LIMIT 5"))
(selstring <- paste0("SELECT st, pwgtp, agep, dis FROM ", tname, " LIMIT 5"))
dbGetQuery(con, selstring )

f <- function(year){
  tname <- paste0("p", year)
  # selstring <- paste0("SELECT st, pwgtp, agep, dis FROM ", tname, " LIMIT 5")
  selstring <- paste0("SELECT st, pwgtp, agep, relp, dis FROM ", tname)
  df <- dbGetQuery(con, selstring ) %>%
    mutate(year=year)
  return(df)
}

system.time(df <- ldply(2010:2018, f))
glimpse(df)
dbDisconnect(con)
count(df, year)
summary(df)

a <- proc.time()
df2 <- df %>%
  mutate(agroup=case_when(agep < 21 ~ "young",
                          agep >= 21 & agep < 65 ~ "workage",
                          agep >= 65 ~ "old",
                          TRUE ~ "error")) %>%
  mutate(stabbr=stcodes$stabbr[match(st, as.integer(stcodes$stfips))],
         igq=(relp==16)) %>%
  group_by(year, stabbr, igq, agroup, dis) %>%
  summarise(n=n(), pop=sum(pwgtp)) %>%
  ungroup
b <- proc.time()
b - a
glimpse(df2)
saveRDS(df2, here::here("data", "age_disb.rds"))

```

```{r ONETIME_lf_age}
#.. BLS Table 14 labor force by age ----
# States:  Employment status of the civilian noninstitutional population by 
#   sex, race, Hispanic or Latino ethnicity, marital status, and detailed age, 2003 annual averages
# download here:  https://www.bls.gov/lau/ex14tables.htm

f <- function(year){
  vnames <- c("stfips", "group_code", "stname", "group", "civnipop", "civlf", "lfpct", "civemp", "emppct", "civunemp", "unemppct")
  y2 <- str_sub(year, 3, 4)
  df <- read_excel(here::here("data/bls", paste0("table14full", y2, ".xlsx")))
  df <- df[-c(1:7), 1:11] %>%
    setNames(vnames) %>%
    mutate(year=year) %>%
    filter(!is.na(stname))
  df
}

df <- ldply(2003:2018, f)
glimpse(df)

count(df, year)
count(df, stfips, stname) # 50 states plus DC
count(df, group_code, group)

df2 <- df %>%
  filter(str_detect(group, "Total,")) %>%
  mutate(stabbr=stcodes$stabbr[match(stfips, stcodes$stfips)]) %>%
  select(year, stabbr, stname, group_code, group, civnipop, civlf, civemp, civunemp) %>%
  mutate_at(vars(c(group_code, civnipop, civlf, civemp, civunemp)), ~as.numeric(.))

glimpse(df2)
ht(df2)
count(df2, stabbr, stname)
count(df2, group_code, group)
saveRDS(df2, here::here("data", "lf_age.rds"))

```


```{r ONETIME_lf_total}
# LAUS Local Area Unemployment Statistics ----
# url <- "https://download.bls.gov/pub/time.series/la/la.data.3.AllStatesS" # SA state data
# urlbase <- "https://download.bls.gov/pub/time.series/la/"
fn <- "la.data.3.AllStatesS" # SA state data
download.file(paste0(urlbase, fn), here::here("data/bls", fn), mode="wb")

# measure_code	measure_text
# 03	unemployment rate
# 04	unemployment
# 05	employment
# 06	labor force

area <- read_tsv("https://download.bls.gov/pub/time.series/la/la.area") %>%
  filter(area_type_code=="A")
area # 50 states DC PR

df <- read_tsv(here::here("data/bls", fn))
glimpse(df)
# https://download.bls.gov/pub/time.series/la/la.area_type

df2 <- df %>%
  filter(str_sub(series_id, -2, -1)=="06") %>%
  mutate(stfips=str_sub(series_id, 6, 7)) %>%
  left_join(area %>% mutate(stfips=str_sub(area_code, 3, 4)) %>% select(stfips, stname=area_text), by = "stfips") %>%
  mutate(stabbr=get_stabbr(stfips))
ht(df2)
count(df2, period)
count(df2, stfips, stabbr, stname)
df2 %>% filter(stabbr=="NY", year>=2016) # through Dec 2019 (preliminary)

df3 <- df2 %>%
  group_by(stabbr, year) %>%
  summarise(lf=mean(value)) %>%
  ungroup
glimpse(df3)

df3 %>%
  filter(stabbr=="TN") %>%
  ggplot(aes(year, lf)) +
  geom_line() +
  geom_point()

summary(df3) # no NA values
saveRDS(df3, here::here("data", "lf_total.rds"))

# series_id                     	year	period	       value	footnote_codes
# LASST010000000000003          	1976	M01	         6.7	

# area_type_code	area_code	area_text	display_level	selectable	sort_sequence
# A	ST0100000000000	Alabama	0	T	1
# A	ST0200000000000	Alaska	0	T	146

# footnote_code	footnote_text
# A	Area boundaries do not reflect official OMB definitions.
# N	Not available.
# P	Preliminary.
# V	The survey was not conducted due to bad weather. Interpolated data were seasonally adjusted.
# W	The household survey was not conducted for this month due to bad weather. Data were interpolated.
# Y	Data reflect controlling to interpolated statewide totals because the survey was not conducted.

```


# Get previously prepared data
```{r pop}
#.. Population ----
pop <- readRDS(here::here("data", "popagesex.rds")) %>%
  filter(vtype=="pop", src=="est", sex==0, month==7) %>%
  select(stabbr, year, age, value)
glimpse(pop)
pop %>%
  filter(stabbr=="NY", age==999)
pop %>%
  filter(stabbr=="NY", year==2010)
```

```{r usmdb}
usmdb <- readRDS(file=here::here("data", "usmdb.rds"))
precis2(usmdb)
# %>% filter(age==65)

yrs <- c(2000, 2017)
usmdb %>%
  filter(age==65, year %in% yrs) %>%
  select(stabbr, year, ex) %>%
  pivot_wider(names_from = year, values_from = ex) %>%
  # mutate(diff=`2017` - `1959`) %>%
  mutate(diff=.[[3]] - .[[2]]) %>%
  arrange(-diff)

sts <- c("NY", "HI", "OK", "MS")
atage <- 65
usmdb %>%
  filter(age==atage) %>%
  filter(stabbr %in% sts) %>%
  ggplot(aes(year, ex, colour=stabbr)) +
  geom_line() +
  geom_point() +
  ggtitle(paste0("Life expectancy at age ", atage, ", selected states")) +
  scale_x_continuous(name=NULL, breaks=seq(1950, 2030, 5)) +
  scale_y_continuous(name="Life expectancy", breaks=seq(0, 30, 1)) +
  theme_bw()

```

```{r acsdb}
# ACS disability data
acsdb <- readRDS(here::here("data", "age_disb.rds"))
glimpse(acsdb)
precis2(acsdb)
# get total rates of disability
# igq is institutionalized group quarters

acsdb %>%
  filter(!igq) %>%
  group_by(year, stabbr) %>%
  summarise(n=sum(n), disb_pop=sum(pop * (dis==1)), pop=sum(pop)) %>%
  mutate(disb_rate=disb_pop / pop * 100) %>%
  filter(stabbr=="AL") 

```

```{r lf_age}
# adult dependency ratio
lf_age <- readRDS(here::here("data", "lf_age.rds"))
precis2(lf_age)
count(lf_age, group_code, group)

df2 <- lf_age %>%
  mutate(youngold=ifelse(group_code <25, "young", "old")) %>%
  group_by(year, stabbr, youngold) %>%
  summarise_at(vars(civnipop, civlf, civemp), ~sum(., na.rm=TRUE)) %>%
  group_by(year, stabbr) %>%
  mutate(nilf=civnipop - civlf,
         totpop=sum(civnipop), # total population of all age groups in the data (16+)
         totlf=sum(civlf), # total civilian labor force of all age groups in the data (16+)
         adr=nilf / totlf * 100) %>% # dependends (nilf) of each age group vs total labor force
  ungroup
df2

# which states had big changes in the adr for 65+

```

```{r lf_total}
lftot <- readRDS(here::here("data", "lf_total.rds"))
```



# OLD STUFF BELOW HERE
## National data
```{r census_us, include=FALSE}
# projections
# https://www.census.gov/data/datasets/2017/demo/popproj/2017-popproj.html

# age sex race hisp origin
# https://www2.census.gov/programs-surveys/popproj/datasets/2017/2017-popproj/np2017_d1.csv
df <- read_csv("https://www2.census.gov/programs-surveys/popproj/datasets/2017/2017-popproj/np2017_d1.csv")
glimpse(df)

# births
# https://www2.census.gov/programs-surveys/popproj/datasets/2017/2017-popproj/np2017_d2.csv

# deaths
# https://www2.census.gov/programs-surveys/popproj/datasets/2017/2017-popproj/np2017_d3.csv


```

## State data
```{r census_state, include=FALSE}


```



