---
title: "Revenue Indicator Analysis"
author: "Don Boyd"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_notebook:
    df_print: paged
    toc: yes
    toc_depth: 5
always_allow_html: yes    
    
editor_options: 
  chunk_output_type: console
---

<!-- comments -->
<!-- -->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


```{r system_specific_values, include=FALSE}

```


```{r globals, include=FALSE}
# 11 states with diverse economies and tax structures
sts <- c("CA", "CT", "FL", "IL", "NJ", "NY", "PA", "TN", "TX", "WA", "WY")
state.name[match(sts, state.abb)]
stus <- c("US", sts)

njcomps <- c("CT", "DE", "MD", "NY", "PA")

ilcomps <- c("IL", "US", "IN", "IA", "KY", "MI", "MO", "WI")

recession_start <- as.Date("2007-12-01")
recession_end <- as.Date("2009-06-01")

# colors lighter to darker
# #fee0d2 #fc9272 #de2d26
red2 <- c("#fee0d2", "#de2d26")
# #deebf7 #9ecae1 #3182bd
blue2 <- c("#deebf7", "#3182bd")
# #e5f5e0 #a1d99b #31a354
green2 <- c("#e5f5e0", "#31a354")

```


```{r libraries, include=FALSE}
library("magrittr")
library("plyr") # needed for ldply; must be loaded BEFORE dplyr
library("tidyverse")
options(tibble.print_max = 60, tibble.print_min = 60) # if more than 60 rows, print 60 - enough for states
# ggplot2 tibble tidyr readr purrr dplyr stringr forcats

library("scales")
library("hms") # hms, for times
library("lubridate") # lubridate, for date/times
library("vctrs")

library("grDevices")
library("knitr")
library("kableExtra")

library("btools")
library("bdata")
library("BEAData")
library("qtax")

library("DT") # for datatable

library("zoo") # for rollapply

library("broom") # for automating the cleanup of complex output

library("quantmod")
library("forecast")
library("tsoutliers")
library("timetk")

library("janitor")

library("ggrepel")

library("maps")
library("usmap")
library("gridExtra")

```


```{r functions}
ma_db <- function(x, period) {
  # create trailing moving average of x of length period
  zoo::rollapply(x, period, function(x) mean(x, na.rm=TRUE), fill=NA, align="right")
}

```


```{r map_setup}
#.. Functions ----
theme_map <- function(base_size=9, base_family="") {
  # see:
  # https://socviz.co/maps.html
  # https://github.com/kjhealy/socviz
  # https://cran.r-project.org/web/packages/usmap/vignettes/mapping.html
  require(grid)
  theme_bw(base_size=base_size, base_family=base_family) %+replace%
    theme(axis.line=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank(),
          axis.title=element_blank(),
          panel.background=element_blank(),
          panel.border=element_blank(),
          panel.grid=element_blank(),
          panel.spacing=unit(0, "lines"),
          plot.background=element_blank(),
          legend.justification = c(0, 0),
          legend.position = c(0, 0)
    )
}


statemap <- function(mapdf, mapvar, cutpts, cutlabels, map_colors, map_title, legend_title){
  mapdf <- mapdf %>%
    rename(mapvalue=mapvar) %>%
    mutate(data_group=cut(mapvalue, cutpts, labels=cutlabels),
           mapvarname=mapvar) %>%
    select(stabbr, mapvarname, mapvalue, data_group)
  
  mdata <- left_join(
    usmap::us_map() %>% arrange(full, piece, order), 
    mapdf %>% rename(abbr=stabbr),
    by="abbr")
  
  p <- mdata %>%
    ggplot(aes(x = x, y = y)) +
    geom_polygon(aes(fill=data_group, group = group),
               color = "gray90", size = 0.1) +
    scale_fill_manual(values=map_colors, drop=TRUE, na.translate=FALSE) + # na.translate drops NA from the legend
    coord_equal() + 
    guides(fill=guide_legend(title=legend_title)) +
    ggtitle(map_title) +
    geom_text(data = bdata::statemap_labels %>% filter(stabbr != "DC"), aes(x, y, label = stabbr), size = 3) +
    theme_map() +
    theme(legend.position = "right") +
    theme(plot.title = element_text(size = 14, face = "bold"))
  
  #   labs(caption=capt) +
  #   theme(plot.caption = element_text(hjust=0, size=8))
  return(p)
}

```


# Introduction
Goals:

* Drive meaningful, actionable insights for policymakers
* Long-term perspective
* Facilitate legitimate state comparisons
* Additional relevant insights (beyond those in current measure)
* Robust, accurate, and timely data sources

# Data preparation
```{r ONETIME_gdppi, eval=FALSE}
gdppi.q <- getSymbols("GDPCTPI", src="FRED", auto.assign=FALSE) %>%
  tk_tbl(rename_index = "date")
gdppi.q %>% ht
gdppi.q <- gdppi.q %>%
  rename(gdppi=2) %>%
  mutate(freq="q") %>%
  mutate(year=year(date))

gdppi.a <- getSymbols("A191RG3A086NBEA", src="FRED", auto.assign=FALSE) %>%
  tk_tbl(rename_index = "date") %>%
  rename(gdppi=A191RG3A086NBEA) %>%
  mutate(freq="a", year=year(date))

gdppi <- bind_rows(gdppi.q, gdppi.a) %>%
  select(year, date, gdppi, freq)
ht(gdppi)

gdppi %>% ggplot(aes(date, gdppi, colour=freq)) + geom_line() + geom_point()
gdppi %>% 
  filter(year >= 2000) %>% 
  ggplot(aes(date, gdppi, colour=freq)) + 
  geom_line() + 
  geom_point() +
  scale_x_date(name=NULL, breaks=seq(as.Date("2000-01-01"), as.Date("2030-01-01"), by="2 years"), date_labels="%Y")

gdppi
saveRDS(gdppi, here::here("data", "gdppi.rds"))

```


```{r ONETIME_qdata, eval=FALSE}
gdppiq <- readRDS(here::here("data", "gdppi.rds")) %>%
  filter(freq=="q") %>%
  select(date, gdppi)
ht(gdppiq)
data(qtax)
glimpse(qtax)
glimpse(sgdp.q)
sgdp.q %>% filter(stabbr=="US") %>% ht

# prepare sgdp -- drop real, compute us
sgdpq2 <- sgdp.q %>%
  select(stabbr, date, gdp) %>%
  group_by(date) %>%
  mutate(gdp=ifelse(stabbr=="US", sum(gdp[stabbr != "US"]), gdp)) %>%
  ungroup
sgdpq2 %>% filter(stabbr=="US") %>% ht

latestq <- "2019-07-01"
rqtax <- qtax %>%
  filter(vname=="tottax") %>%
  select(stabbr, date, tax=value) %>%
  left_join(gdppiq, by="date") %>%
  left_join(sgdpq2, by = c("stabbr", "date")) %>%
  mutate(tax=tax / 1000, # billions of dollars
         rtax=tax * gdppiq$gdppi[gdppiq$date==latestq] / gdppi,
         rgdp=gdp * gdppiq$gdppi[gdppiq$date==latestq] / gdppi) %>%
  select(-gdppi)
summary(rqtax) # only gdp and rgdp have NAs
saveRDS(rqtax, here::here("data", "rqtax.rds"))

```


```{r ONETIME_adata, eval=FALSE}
# annual
# count(slgfin, aggvar) %>% filter(str_detect(aggvar, "tax"))

gdppia <- readRDS(here::here("data", "gdppi.rds")) %>%
  filter(freq=="a") %>%
  select(year, gdppi)

# compute US gdp (sum of states)
sgdp.a %>% filter(stabbr=="US") # 1997+
sgdpa2 <- sgdp.a %>%
  select(year, stabbr, gdp) %>%
  group_by(year) %>% 
  mutate(gdp=ifelse(stabbr=="US", sum(gdp[stabbr!="US"]), gdp)) %>%
  ungroup
ht(sgdpa2)
sgdpa2 %>% filter(stabbr=="US") # 1997+

latesty <- 2017
rsltax <- slgfin %>%
  filter(aggvar=="tottax", level==1, year>=1997) %>% # we don't have gdp before 1997
  select(stabbr, year, tax=value) %>%
  left_join(sgdpa2, by = c("stabbr", "year")) %>%
  left_join(gdppia %>% select(year, gdppi), by = "year") %>%
  mutate(rtax=tax * gdppia$gdppi[gdppia$year==latesty] / gdppi,
         rgdp=gdp * gdppia$gdppi[gdppia$year==latesty] / gdppi) %>%
  select(-gdppi) %>%
  mutate(taxgdp=tax / gdp / 10) %>% # as a percent
  group_by(year) %>%
  mutate(pctus=taxgdp / taxgdp[stabbr=="US"] * 100,
         rank=ifelse(stabbr=="US", NA, rank(-taxgdp))) %>%
  ungroup
glimpse(rsltax)
ht(rsltax)
rsltax %>% filter(stabbr=="NY") %>% tail(20)
rsltax %>% filter(stabbr=="IL") %>% tail(20)
rsltax %>% filter(stabbr=="US") %>% tail(20)

rsltax %>% filter(year==2017) %>% arrange(-taxgdp)
saveRDS(rsltax, here::here("data", "rsltax.rds"))

```


# Revenue growth measures
```{r get_rqtax}
# keep all states in the data until we get to graphing

rqtax <- readRDS(here::here("data", "rqtax.rds"))

df <- rqtax %>%
  group_by(stabbr) %>%
  arrange(date) %>%
  mutate(rtaxma=ma_db(rtax, 4),
         rtaxma_pchya=rtaxma / lag(rtaxma, 4) * 100 - 100,
         rgdpma=ma_db(rgdp, 4)) %>%
  filter(year(date) >= 1980) %>%
  ungroup

glimpse(df)

# function to smooth and index based on a given period
si <- function(df, start){
  df2 <- df %>%
    filter(date >= start) %>%
    group_by(stabbr) %>%
    arrange(date) %>%
    mutate(rn=row_number(),
           rtaxma_fit=loess(rtaxma ~ rn)$fitted,
           pd_rtfit=rtaxma_fit / rtaxma_fit[date==start] * 100 - 100,
           pd_rtma=rtaxma / rtaxma[date==start] * 100 - 100) %>%
    ungroup
  df2
}

```


## Alternative growth measure
## How the level of taxation has changed
### The level
```{r altlevel}
# smooth and index
pdata <- si(df, "2000-01-01") %>%
  filter(stabbr %in% stus) %>%
  mutate(stabbr=factor(stabbr, levels=stus))

p <- pdata %>%
  ggplot() +
  geom_line(aes(date, rtaxma)) +
  geom_rect(data = data.frame(xmin = recession_start, xmax = recession_end,
                              ymin = -Inf, ymax = Inf),
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "grey", alpha = 0.25) +
  stat_smooth(aes(date, rtaxma), method="loess", fill="lightblue") +
  # geom_line(aes(date, rtaxma_fit), colour="red", size=1) +
  facet_wrap(~stabbr, ncol=3, scales="free") +
  scale_y_continuous("Billions of 2019 $", labels=scales::comma) +
  scale_x_date(name=NULL) +
  theme_bw() +
  ggtitle("Inflation-adjusted quarterly state tax revenue, 4-quarter moving average",
          subtitle="Actual values and trend")
p
ggsave(here::here("results", "rqtax_v_trend.png"), plot=p, width=10, height=10, scale=1)  

```


### The level in a single state
```{r}
# single state graph
# each state vs US
idate <- "2010-01-01"
pdata <- si(df, idate) %>%
  # get the US value and fit
  select(stabbr, date, rtaxma, pd_rtma, rtaxma_fit, pd_rtfit) %>%
  pivot_longer(c(rtaxma, pd_rtma, rtaxma_fit, pd_rtfit), names_to= "vname", values_to = "stvalue") %>%
  group_by(date, vname) %>%
  mutate(usvalue=stvalue[stabbr=="US"]) %>%
  pivot_longer(c(stvalue, usvalue), names_to= "govt") %>%
  ungroup

pdata %>%
  filter(vname=="pd_rtfit") %>%
  filter(stabbr %in% stus) %>%
  mutate(stabbr=factor(stabbr, levels=stus)) %>%
  ggplot() +
  geom_line(aes(date, value, colour=govt)) +
  facet_wrap(~stabbr, ncol=3, scales="free") +
  scale_x_date(name=NULL) +
  theme_bw() +
  ggtitle("Inflation-adjusted quarterly state tax revenue, 4-quarter moving average, % change vs. start date",
          subtitle="State and US")

sts <- c("NJ", njcomps)
pdata %>%
  filter(vname=="pd_rtma") %>%
  filter(stabbr %in% sts, govt=="stvalue") %>%
  mutate(stabbr=factor(stabbr, levels=sts)) %>% # to keep in desired order
  ggplot() +
  geom_line(aes(date, value, colour=stabbr)) +
  geom_hline(yintercept = 0) + 
  # scale_y_continuous("Billions of 2019 $", labels=scales::comma) +
  scale_x_date(name=NULL) +
  theme_bw() +
  ggtitle("Inflation-adjusted quarterly state tax revenue, 4-quarter moving average, % change vs. start date",
          subtitle="State and US")
 


# New Jersey and nearby states graph 
fnj <- function(idate){
  sts <- c("NJ", "US", njcomps)
  groupvals <- c("NJ", "comp", "US")
  grouplabs <- c("New Jersey", "Nearby states", "United States")
  pdata <- si(df, idate) %>%
    # get the US value and fit
    select(stabbr, date, pd_rtma) %>%
    pivot_longer(c(pd_rtma), names_to= "vname", values_to = "stvalue") %>%
    group_by(date, vname) %>%
    mutate(usvalue=stvalue[stabbr=="US"]) %>%
    pivot_longer(c(stvalue, usvalue), names_to= "govt") %>%
    ungroup %>%
    filter(vname=="pd_rtma") %>%
    filter(stabbr %in% sts, govt=="stvalue") %>%
    mutate(group=case_when(stabbr %in% c("NJ", "US") ~stabbr,
                           stabbr %in% njcomps ~ "comp",
                           TRUE ~ "other")) %>%
    group_by(group, date) %>%
    summarise(value=mean(value)) %>%
    ungroup %>%
    mutate(group=factor(group, levels=groupvals, labels=grouplabs))
  pdata
}

idate <- "2010-01-01"
p <- fnj(idate)  %>%
  ggplot() +
  geom_line(aes(date, value, colour=group)) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept=as.Date(idate), linetype="dotted") +
  scale_y_continuous("Percent change since start of period", breaks=seq(-20, 50, 5)) +
  scale_x_date(name=NULL) +
  scale_colour_manual(values=c("blue", "red", "black")) +
  ggtitle("State tax revenue: New Jersey, average of nearby states, and the U.S.",
          subtitle="% change from start in inflation-adjusted 4-quarter moving average") +
  labs(caption="Nearby states: Connecticut, Delaware, Maryland, New York & Pennsylvania") +
  theme_bw() +
  theme(legend.title = element_blank()) +
  theme(plot.caption = element_text(hjust=0, size=rel(.8)))
p
ggsave(here::here("results", "njrqtaxma_v_nearby.png"), plot=p, width=10, height=6, scale=1)   

# individual states
sts <- c("NJ", njcomps)
pdata <- si(df, idate) %>%
  filter(stabbr %in% sts) %>%
  select(stabbr, date, pd_rtma)

clrs <- c('red', '#4daf4a', '#984ea3', 'blue', '#ff7f00', 'black')
p <- pdata %>%
  ggplot() +
  geom_line(aes(date, pd_rtma, colour=stabbr, size=stabbr)) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept=as.Date(idate), linetype="dotted") +
  scale_y_continuous("Percent change since start of period", breaks=seq(-20, 50, 5)) +
  scale_x_date(name=NULL) +
  scale_colour_manual(values=clrs) +
  scale_size_manual(values=c(1, 1, 1, 2, 1, 1)*.8) +
  ggtitle("State tax revenue: New Jersey and nearby states",
          subtitle="% change from start in inflation-adjusted 4-quarter moving average") +
  theme_bw() +
  theme(legend.title = element_blank()) +
  theme(plot.caption = element_text(hjust=0, size=rel(.8)))
p
ggsave(here::here("results", "njrqtaxma_v_nearby_idiv.png"), plot=p, width=10, height=6, scale=1) 


idate <- "2010-01-01"
md <- si(df, idate) %>%
  filter(stabbr != "US", date==max(date)) %>%
  select(stabbr, pd_rtma)
quantile(md$pd_rtma)
cutpoints <- c(-Inf, 0, 20, 30, 40, Inf)
cutvals <- cut(md$pd_rtma, cutpoints, right=FALSE)
table(cutvals) %>% as.data.frame()
cutlabs <- c("< 0", "0 to < 20", "20 to < 30", "30 to < 40", "40+")
cbind(cutlabs, table(cutvals) %>% as.data.frame())

clrs <- c(red2 %>% rev, "lightgrey", blue2)

p <- statemap(mapdf=md, mapvar="pd_rtma", 
         cutpts=cutpoints, cutlabels=cutlabs, map_colors=clrs,
         map_title="% change in inflation-adjusted tax revenue from 2010 to most recent quarter",
         legend_title="% change")
p
ggsave(here::here("results", "rqtax_v_2010_map.png"), plot=p, width=16, height=9, scale=.7)  

```


### Map of level - percent above or below trend

```{r map1_pdiff, fig.height=9 fig.width=16}
# map all states
pdata <- df %>%
  filter(year(date) >= 2000) %>%
  mutate(rtaxma=rtaxma / 1000) %>%
  # create smoother virtually (or actually) identical to stat_smooth
  group_by(stabbr) %>%
  arrange(date) %>%
  mutate(rn=row_number(),
         fit=loess(rtaxma ~ rn)$fitted, 
         pdiff=rtaxma / fit * 100 - 100) %>%
  ungroup %>%
  filter(date==max(date), stabbr != "US")

quantile(pdata$pdiff)

cutpts <- c(-Inf, -3, -1, 1, 3, Inf)
(pcuts <- cut(pdata$pdiff, cutpts))
levs <- levels(pcuts)
cutlabels <- c("3% or more below", "1% to 3% below",
          "within 1% of trend",
          "1% to 3% above", "3% or more above")
map_title <- "States with tax revenue above or below longer-term trend"
legend_title <- "% above or below trend"

p <- statemap(mapdf=pdata, 
              mapvar="pdiff", 
              cutpts=cutpts, cutlabels=cutlabels, map_colors=c(red2 %>% rev, "#f0f0f0", green2),
              map_title=map_title, legend_title=legend_title)
p

ggsave(here::here("results", "rqtax_v_trend_map.png"), plot=p, width=16, height=9, scale=.7)  

```


## Short-term growth
```{r altgrow}
# shorter term growth measure
p <- df %>%
  filter(year(date) >= 2010,
         stabbr %in% stus) %>%
  mutate(stabbr=factor(stabbr, levels=stus)) %>%
  
  # create smoother virtually (or actually) identical to stat_smooth
  group_by(stabbr) %>%
  arrange(date) %>%
  mutate(rn=row_number(),
         fit=loess(rtaxma_pchya ~ rn)$fitted) %>%
  ungroup %>%
  group_by(date) %>% 
  mutate(usfit=fit[stabbr=="US"]) %>%
  ungroup %>%
  
  ggplot(aes(date, rtaxma_pchya)) +
  geom_line() +
  stat_smooth(method="loess", fill="lightblue") +
  geom_line(aes(date, usfit), colour="red", size=1) +
  geom_hline(yintercept = 0) +
  facet_wrap(~stabbr, ncol=3, scales="free") +
  scale_y_continuous("% change vs. year ago") +
  scale_x_date(name=NULL) +
  theme_bw() +
  ggtitle("% change vs. prior year in 4-quarter average of inflation-adjusted state tax revenue",
          subtitle="Actual % change and trend; red line is U.S. trend")
p
ggsave(here::here("results", "rqtax_growth_v_trend.png"), plot=p, width=10, height=10, scale=1)  

```




```{r}
# break states into 3 groups: slower than US for last 4 quarters, faster, and neither
pdata <- xtrap3 %>%
  select(stabbr, date, vname, state=rvalmapchya) %>%
  group_by(vname, date) %>%
  mutate(us=state[stabbr=="US"],
         gtus=state > us,
         ltus=state < us) %>%
  group_by(stabbr, vname) %>%
  arrange(date) %>%
  mutate(gt4=ma_db(gtus, 4) * 4,
         lt4=ma_db(ltus, 4) * 4)

# get state groups
stgroups <- pdata %>%
  filter(date==max(date)) %>%
  mutate(stgroup=case_when(gt4 == 4 ~ "faster",
                           lt4 == 4 ~ "slower",
                           TRUE ~ "mixed")) %>%
  select(stabbr, date, vname, stgroup, gt4, lt4, stgroup) %>%
  ungroup
stgroups
count(stgroups, stgroup)  
stgroups %>% filter(stgroup=="faster")
stgroups %>% filter(stgroup=="slower")

slow <- stgroups %>% filter(stgroup=="slower") %>% .$stabbr
fast <- stgroups %>% filter(stgroup=="faster") %>% .$stabbr
mixed <- stgroups %>% filter(stgroup=="mixed") %>% .$stabbr

pdata %>%
  select(stabbr, date, vname, state, us) %>%
  filter(year(date) >= 2014, stabbr %in% mixed) %>%
  pivot_longer(c(state, us)) %>%
  ggplot(aes(date, value, colour=name)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  facet_wrap(~stabbr, ncol=4, scales = "fixed") +
  # scale_x_date(breaks="3 months") +
  scale_y_continuous(breaks=seq(-30, 30, 2.5)) +
  theme_bw()

```


## Revenue growth and economy
```{r}
glimpse(df)

# start <- "2010-01-01"
start <- "2010-01-01"
# sts <- c("CA", "NY", "IL", "US")
# sts <- c("FL", "TX", "PA", "US")
# sts <- c("FL", "TX", "PA", "ME", "US")
# 
# sts <- c("FL", "TX", "IL", "CA")
df2 <- df %>%
  filter(date >= start) %>%
  group_by(stabbr) %>%
  mutate(irgdpma=rgdpma / rgdpma[date==start] * 100 - 100,
         irtaxma=rtaxma / rtaxma[date==start] * 100 - 100,
         diff=irtaxma - irgdpma) %>%
  ungroup

gaps <- df2 %>% filter(date==max(date)) %>% arrange(diff)
xstates <- c("AK", "DC", "US")
top <- gaps %>% filter(!stabbr %in% xstates) %>% top_n(n=4, wt=diff) %>% arrange(-diff)
bottom <- gaps %>% filter(!stabbr %in% xstates) %>% top_n(n=4, wt=-diff) %>% arrange(-diff)

(sts2 <- c(top$stabbr, "US", bottom$stabbr))

# df2 %>% 
#   filter(stabbr %in% sts) %>%
#   select(stabbr, date, irgdpma, irtaxma) %>%
#   pivot_longer(c(irgdpma, irtaxma)) %>%
#   mutate(stabbr=factor(stabbr, levels=sts)) %>%
#   ggplot(aes(date, value, colour=name)) +
#   geom_line() +
#   geom_point() +
#   geom_hline(yintercept = 0) +
#   scale_y_continuous(breaks=seq(-50, 50, 10)) +
#   facet_wrap(~stabbr, ncol=3)
# 
# comps <- c("WI", "IA", "IN", "KY", "MO", "MI")
# sts2 <- c("IL", "US", "IN", "IA", "KY", "MI", "MO", "WI")
# df2 %>% 
#   filter(stabbr %in% sts) %>%
#   select(stabbr, date, irgdpma, irtaxma) %>%
#   pivot_longer(c(irgdpma, irtaxma)) %>%
#   mutate(stabbr=factor(stabbr, levels=sts)) %>%
#   ggplot(aes(date, value, colour=name)) +
#   geom_line() +
#   geom_point() +
#   geom_hline(yintercept = 0) +
#   scale_y_continuous(breaks=seq(-50, 50, 10)) +
#   facet_wrap(~stabbr, ncol=3)

pdata <- df2 %>%
  filter(stabbr %in% stus) %>%
  select(stabbr, date, starts_with(("ir"))) %>%
  pivot_longer(starts_with("ir")) %>%
  mutate(stabbr=factor(stabbr, levels = stus))

p <- pdata %>%
  mutate(name=factor(name, 
                     levels=c("irtaxma", "irgdpma"),
                     labels=c("Tax revenue", "State GDP"))) %>%
  ggplot() +
  geom_line(aes(date, value, colour=name), size=1) +
  geom_hline(yintercept = 0) +
  facet_wrap(~stabbr, ncol=3, scales="free") +
  scale_y_continuous("Cumulative % change since 2010") + # , breaks=seq(-5, 30, 5) +
  scale_x_date(name=NULL) +
  ggtitle("State tax revenue and GDP growth since 2010",
          subtitle="4-quarter moving average of inflation-adjusted values") +
  theme_bw() +
  theme(legend.position = "right") +
  theme(plot.title = element_text(size = 14, face = "bold")) +
  guides(colour=guide_legend(title=NULL))
p
ggsave(here::here("results", "rqtax_gdp.png"), plot=p, width=10, height=10, scale=1)  

```


# Revenue burden measures
```{r}
rsltax <- readRDS(here::here("data", "rsltax.rds"))
glimpse(rsltax)

pdata <- rsltax %>%
  filter(stabbr %in% stus, year>=2000) %>%
  mutate(stabbr=factor(stabbr, levels = stus)) %>%
  select(stabbr, year, taxgdp)

p <- pdata %>%
  ggplot() +
  geom_line(aes(year, taxgdp)) +
  stat_smooth(aes(year, taxgdp), method="loess", fill="lightblue") +
  # geom_line(aes(date, rtaxma_fit), colour="red", size=1) +
  geom_rect(data = data.frame(xmin = 2007.75, xmax = 2009.5,
                              ymin = -Inf, ymax = Inf),
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "grey", alpha = 0.25) +
  facet_wrap(~stabbr, ncol=3, scales="free") +
  scale_y_continuous("State-local tax revenue as % of GDP") +
  scale_x_continuous(name=NULL, breaks=seq(2000, 2020, 5)) +
  theme_bw() +
  ggtitle("State-local tax revenue as % of GDP",
          subtitle="Actual values and trend")
p
ggsave(here::here("results", "rsltax_gdp_trend.png"), plot=p, width=10, height=10, scale=1)  


p <- pdata %>%
  group_by(year) %>%
  mutate(itaxgdp=taxgdp / taxgdp[stabbr=="US"] * 100 -100) %>%
  ungroup %>%
  filter(stabbr != "US") %>%
  ggplot() +
  geom_line(aes(year, itaxgdp)) +
  stat_smooth(aes(year, itaxgdp), method="loess", fill="lightblue") +
  geom_hline(yintercept = 0, colour="red") +
  # geom_line(aes(date, rtaxma_fit), colour="red", size=1) +
  geom_rect(data = data.frame(xmin = 2007.75, xmax = 2009.5,
                              ymin = -Inf, ymax = Inf),
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "grey", alpha = 0.25) +
  facet_wrap(~stabbr, ncol=3, scales="free") +
  scale_y_continuous("State-local tax revenue as % of GDP") +
  scale_x_continuous(name=NULL, breaks=seq(2000, 2020, 5)) +
  theme_bw() +
  ggtitle("State-local tax revenue as % of GDP, % above or below the US",
          subtitle="Actual values and trend")
p
ggsave(here::here("results", "index_rsltax_gdp_trend.png"), plot=p, width=10, height=10, scale=1)  

p <- rsltax %>%
  filter(stabbr %in% stus) %>%
  mutate(stabbr=factor(stabbr, levels = stus)) %>%
  select(stabbr, year, taxgdp) %>%
  group_by(year) %>%
  mutate(itaxgdp=taxgdp / taxgdp[stabbr=="US"] * 100 -100) %>%
  group_by(stabbr) %>%
  arrange(year) %>%
  mutate(itaxgdpma=ma_db(itaxgdp, 3)) %>%
  ungroup %>%
  filter(stabbr != "US", year>=2000) %>%
  ggplot() +
  geom_line(aes(year, itaxgdp)) +
  geom_line(aes(year, itaxgdpma), colour="darkgreen", size=1) +
  stat_smooth(aes(year, itaxgdp), method="loess", fill="lightblue") +
  geom_hline(yintercept = 0, colour="red") +
  # geom_line(aes(date, rtaxma_fit), colour="red", size=1) +
  geom_rect(data = data.frame(xmin = 2007.75, xmax = 2009.5,
                              ymin = -Inf, ymax = Inf),
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "grey", alpha = 0.25) +
  facet_wrap(~stabbr, ncol=3, scales="free") +
  scale_y_continuous("State-local tax revenue as % of GDP") +
  scale_x_continuous(name=NULL, breaks=seq(2000, 2020, 5)) +
  theme_bw() +
  ggtitle("State-local tax revenue as % of GDP, % above or below the US",
          subtitle="Actual values (black), trend (blue), and 3-year moving average (green)")
p
ggsave(here::here("results", "index_rsltax_gdp_trend_ma.png"), plot=p, width=10, height=10, scale=1)  



ilsts <- c("IL", "WI", "IA", "IN", "KY", "MO")

rsltax %>%
  filter(stabbr %in% ilsts) %>%
  # filter(year %in% seq(2000, 2020, 5)) %>%
  filter(year %in% c(2000, 2005, 2010:2017)) %>%
  ggplot(aes(year, pctus, colour=stabbr)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 100)

rsltax %>%
  filter(stabbr %in% c("CA", "IL", "IN", "NY")) %>%
  ggplot(aes(year, rank, colour=stabbr)) +
  geom_line() +
  geom_point() +
  scale_y_reverse()

ilsts <- c("IL", "WI", "IN", "KY", "MO")
rsltax %>%
  filter(stabbr %in% ilsts) %>%
  filter(year %in% c(2000, 2005, 2010, 2016)) %>%
  mutate(pctus=pctus - 100) %>%
  ggplot(aes(year, pctus, fill=stabbr)) +
  geom_col(position="dodge") +
  geom_hline(yintercept = 0) + 
  theme_bw()

rsltax %>%
  filter(stabbr %in% c("IL", "US")) %>%
  ggplot(aes(year, taxgdp, colour=stabbr)) +
  geom_line() +
  theme_bw()

ilcomps <- c("WI", "IA", "IN", "KY", "MO", "MI")
rsltax %>%
  mutate(grp=case_when(stabbr=="IL" ~ "IL",
                       stabbr=="US" ~ "US",
                       stabbr %in% comps ~ "comp",
                       TRUE ~ "other")) %>%
  # filter(grp!="other") %>%
  filter(grp!="US") %>%
  group_by(year, grp) %>%
  summarise(taxgdp=median(taxgdp)) %>%
  group_by(grp) %>%
  mutate(taxgdp=ma_db(taxgdp, 3)) %>%
  ggplot(aes(year, taxgdp, colour=grp)) +
  geom_line() +
  geom_point() +
  # scale_y_continuous(limits=c(0, NA)) +
  theme_bw()

# Illinois change in pct
p <- rsltax %>%
  mutate(grp=case_when(stabbr=="IL" ~ "IL",
                       stabbr=="US" ~ "US",
                       stabbr %in% comps ~ "comp",
                       TRUE ~ "other")) %>%
  mutate(grp=factor(grp,
                    levels=c("IL", "comp", "other", "US"),
                    labels=c("Illinois", "Nearby states", "All other states", "US average"))) %>% 
  filter(grp!="US") %>%
  group_by(year, grp) %>%
  summarise(taxgdp=median(taxgdp)) %>%
  group_by(grp) %>%
  mutate(diff=taxgdp - taxgdp[year==2000],
         taxgdp2=ma_db(taxgdp, 3),
         diff2=taxgdp2 - taxgdp2[year==2000]) %>%
  filter(year>=2000) %>%
  ggplot(aes(year, diff2, colour=grp)) +
  scale_colour_manual(values=c("black", "darkgreen", "purple", "blue")) +
  geom_line(aes(size=grp)) +
  scale_size_manual(values=c(1.5, 1, 1, 1)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_x_continuous(name=NULL) +
  scale_y_continuous(name="Change since 2000 as % of state GDP",
                     breaks=seq(-2, 2, .25)) +
  guides(colour=guide_legend(title=NULL),
         size=guide_legend(title=NULL)) +
  labs(caption="Nearby states: Indiana, Iowa, Kentucky, Michigan, Missouri, & Wisconsin") +
  ggtitle("State & local taxes as % of GDP: Illinois, median of nearby states, and median of other states",
          subtitle="Change since 2000 in 3-year moving average of taxes as % of state GDP") +
  theme_bw() +
  theme(plot.caption = element_text(hjust=0, size=rel(.8)))
p
ggsave(here::here("results", "ILplot.png"), plot=p, width=10, height=10, scale=1)  

# Illinois above or below us
il <- rsltax %>% filter(stabbr %in% c("IL", "US"))
p <- rsltax %>%
  mutate(grp=case_when(stabbr=="IL" ~ "IL",
                       stabbr=="US" ~ "US",
                       stabbr %in% comps ~ "comp",
                       TRUE ~ "other")) %>%
  mutate(grp=factor(grp,
                    levels=c("IL", "comp", "other", "US"),
                    labels=c("Illinois", "Nearby states", "All other states", "US average"))) %>% 
  filter(stabbr!="US") %>%
  group_by(year, grp) %>%
  summarise(pctus=median(pctus)) %>%
  mutate(diffus=pctus - 100) %>%
  filter(year>=2000) %>%
  ggplot(aes(year, diffus, colour=grp)) +
  scale_colour_manual(values=c("black", "darkgreen", "purple", "blue")) +
  geom_line(aes(size=grp)) +
  scale_size_manual(values=c(1.5, 1, 1, 1)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_x_continuous(name=NULL) +
  # scale_y_continuous(name="% of state GDP",
  #                    breaks=seq(-2, 2, .25)) +
  guides(colour=guide_legend(title=NULL),
         size=guide_legend(title=NULL)) +
  labs(caption="Nearby states: Indiana, Iowa, Kentucky, Michigan, Missouri, & Wisconsin") +
  ggtitle("State & local taxes as % of GDP, minus U.S.: Illinois, median of nearby states, and median of other states",
          subtitle="3-year moving average of taxes as % of state GDP") +
  theme_bw() +
  theme(plot.caption = element_text(hjust=0, size=rel(.8)))
p
ggsave(here::here("results", "ILplot_vsUS.png"), plot=p, width=10, height=10, scale=1)  

#  theme(legend.position = "none")
# pct gdp
top25pct <- rsltax %>%
  filter(stabbr != "US") %>%
  group_by(year) %>%
  summarise(p25=p25(taxgdp),
         p75=p75(taxgdp)) %>%
  pivot_longer(c(p25, p75), values_to = "taxgdp")

stgroups <- rsltax %>%
  mutate(name=case_when(stabbr=="IL" ~ "IL",
                       stabbr=="US" ~ "US",
                       stabbr %in% comps ~ "comp",
                       TRUE ~ "other")) %>%
  group_by(year, name) %>%
  # summarise(taxgdp=median(taxgdp)) %>%
  summarise(taxgdp=sum(value) / sum(gdp) / 10)
  ungroup

pdata <- bind_rows(stgroups, top25pct) %>%
  filter(name %in% grps) %>%
  group_by(name) %>%
  arrange(year) %>%
  mutate(taxgdp2=ma_db(taxgdp, 3)) %>%
  ungroup

grps <- c("IL", "comp", "p75", "US")
labs <- c("Illinois", "Nearby states", "Lowest top-quartile state", "US average")

capt1 <- "Nearby states: Indiana, Iowa, Kentucky, Michigan, Missouri, & Wisconsin"
capt2 <- "Calculation for nearby states: sum of their taxes as % of sum of their GDP."
capt <- paste0(capt1, "\n", capt2)
p <- pdata %>%
  filter(year>=2000) %>%
  mutate(grp=factor(name, levels=grps, labels=labs)) %>%
  ggplot(aes(year, taxgdp2, colour=grp)) +
  scale_colour_manual(values=c("black", "darkgreen", "purple", "blue")) +
  geom_line(aes(size=grp)) +
  scale_size_manual(values=c(1.5, 1, 1, 1)) +
  geom_point() +
  # geom_hline(yintercept = 0) +
  scale_x_continuous(name=NULL) +
  scale_y_continuous(name="% of state GDP",
                    breaks=seq(0, 20, .25)) +
  guides(colour=guide_legend(title=NULL),
         size=guide_legend(title=NULL)) +
  labs(caption=capt) +
  ggtitle("State & local taxes as % of GDP",
          subtitle="3-year moving average") +
  theme_bw() +
  theme(plot.caption = element_text(hjust=0, size=rel(.8)))
p
p1 <- p + 
  theme(legend.direction = 'vertical',
        legend.position = 'right',
        legend.key = element_rect(size = 5),
        legend.key.size = unit(1.5, 'lines')) # put space between legend lines of text

ggsave(here::here("results", "ILplot_taxgdp.png"), plot=p1, width=10, height=6, scale=1) 

rsltax %>% filter(year==2017) %>% arrange(-taxgdp)
```

